Exploring the Role of Feedback in Machine Learning Jupyter Notebooks

### 2024-06-28 report/emse24.tex
refactor(report): fix remaining merge conflicts

In software engineering, tests are employed to document the expected behavior of code. Failing tests indicate when these expectations are no longer met. Similarly, assertions in Jupyter notebooks can serve as documentation for the decisions made during model development. Embedding assertions in the notebook can ensure that any deviations from expected data conditions or model performance are promptly detected, thus maintaining the integrity of the ML pipeline.

@@

Of the total 298K notebooks collected in this study, 25K (8.5\%) contained at least one assertion. From the 8.5\% notebooks, we collected 89.6K assertions. 4.8K assertions (5\%) used methods from external testing libraries. Majority of the assertions however (85K or 95\%) were written using the \lstinline{assert} statement provided by the built-in Python standard library. We removed duplicate assertions for the remainder of the analysis, thus resulting in 27.1K assertions.

@@

@@

Existence checks are a fundamental component of data preprocessing in ML, serving multiple critical purposes to ensure data integrity before analysis and modeling. These checks primarily focus on verifying the presence of necessary columns and the absence of missing values within those columns. Such validations are especially crucial after data preprocessing steps, where transformations might inadvertently introduce null values. These checks are also crucial for maintaining the accuracy and efficiency of data handling processes, ensuring that the datasets are ready for robust machine learning applications.

@@

Assertion \emph{A3} in Table~\ref{tab:maths-check} checks that the dimension differences between height of input and the filter, when divided by the stride, result in an integer, ensuring that convolution operations are dimensionally consistent. \emph{A25} monitors the standard deviation of outputs to decide the appropriateness of using batch normalisation, thus guiding the model optimization process based on statistical behavior. \emph{A56} verifies the uniformity of certain computations, essential for maintaining consistency in state or parameter arrays over iterations. Similarly \emph{A64} checks for the symmetry of a matrix, a crucial property in many algorithms, particularly those involving covariance matrices and distance calculations.

When working with neural networks and other statistical models, it is imperative to ensure that mathematical properties are preserved throughout operations involving arrays and matrices. This rigor is captured through the assertions in Table~\ref{tab:maths-check} tha validate these properties after matrix manipulations. These assertions are not just precautionary but are vital for confirming that the underpinnings of ML algorithms align with expected mathematical principles, thereby ensuring the robustness and reliability of model computations and outcomes.

@@

% TODO: verify if this applies to recommender systems as well?
For example, \emph{A7} in Table~\ref{tab:model-perf-explicit} checks for the expected number of neighbors as an indirect measure of model performance in specific scenarios like recommendation systems or clustering. Meanwhile, direct comparisons of model outputs with expected results, as seen in \emph{A15} validate the model's predictive accuracy and reliability under operational conditions. Other assertions, such as \emph{A19}, ensure that the learned parameters align closely with theoretically or empirically derived values, further cementing the model’s statistical validity. Range checks like \emph{A38} confirm that the performance metrics such as \emph{AUC} and \emph{Gini coefficient} fall within acceptable boundaries, indicating good predictive performance without overfitting.

@@

Assertion \emph{A75} ensures that the regularization parameter \lstinline{reg} is either 'none' or 'l2'. This check is important for validating that the specified regularization method is among the acceptable options, thereby preventing potential misconfigurations that could lead to unexpected behavior or suboptimal model performance.

@@

Type check assertions are essential for maintaining data integrity, ensuring compatibility between different components of the machine learning pipeline, and preventing errors and unexpected behavior. By verifying the types of input features, models, data structures, and other objects, these assertions contribute to the overall reliability and robustness of the machine learning workflow.

@@

The visualisation in VA3 presents a two-dimensional scatter plot showing the decision boundary of a Support Vector Machine (SVM) model, with red and yellow points representing two different classes and a background color gradient indicating the decision function of the SVM. This visualisation aids in understanding the model's ability to distinguish between classes and the distance of the samples from the decision boundary. The visualisations in VA28 display the training loss and development recall, respectively. The left graph shows a decrease in training loss over time, indicating learning progress, while the right graph shows development recall over time, highlighting improvements in the model's ability to correctly identify positive instances. The visualisations in VA11 depict key metrics for evaluating the performance of a machine learning model in text generation or machine translation. The left plot shows a steady increase in the development BLEU score over the training iterations, indicating improving quality of the model's output, while the right plot displays a steady decrease in training loss, suggesting effective learning and error minimization. In VA107, the scatter plot displays the predicted versus true values of Nitrogen Dioxide (NO2) levels, used to assess the model's performance in predicting continuous variables.

>>>>

In software engineering, tests are employed to document the expected behaviour of code. Failing tests indicate when these expectations are no longer met. Similarly, assertions in Jupyter notebooks can serve as documentation for the decisions made during model development. Embedding assertions in the notebook can ensure that any deviations from expected data conditions or model performance are promptly detected, thus maintaining the integrity of the ML pipeline.

@@

25K (8.5\%) notebooks contained at least one assertion. From the 8.5\% notebooks, we collected 89.6K assertions. 4.8K assertions (5\%) used methods from external testing libraries. Majority of the assertions however (85K or 95\%) were written using the \lstinline{assert} statement provided by the built-in Python standard library. We removed duplicate assertions for the remainder of the analysis, thus resulting in 27.1K assertions.

@@

\begin{figure}
  \centering
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{common-print-nodes.pdf}
    \caption{Most common AST nodes inside print statements}
    \label{fig:common-print-nodes}
  \end{minipage}
  \hfill
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{common-print-constant-types.pdf}
    \caption{Most common type of constant inside print statements}
    \label{fig:common-print-constant-types}
  \end{minipage}
\end{figure}

We collected 1.4M print statements from 180K notebooks (60\% of the total notebooks analysed in this study). Unlike assert statements, print statements do not have a fixed lexical semantic since the Python ast module interprets them as regular function calls. Additionally, print statements can be written in several forms---the \lstinline{print} function may have multiple arguments, any form of string interpolation supported by Python and output from function calls. Figure~\ref{fig:common-print-nodes} shows that the most common nodes appear inside of print statements are constants followed by function calls. Figure~\ref{fig:common-print-constant-types} further validates that the constants are almost entirely of type string. Based on these observations, we disect the print statements into two components---text written in natural language (from the \emph{Constant} nodes) and code (from all other nodes). Figure~\ref{fig:common-print-constants} and Figure~\ref{fig:common-print-not-constants} present the most common bi-grams that appear in the natural text and code of print statements respectively. Both figures collectively indicate that print statements are frequently used to print various performance metrics of ML models.

\begin{figure}
  \centering
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-print-constants.pdf}
    \caption{Most common bi-grams that appear in natural text inside of print statements.}
    \label{fig:common-print-constants}
  \end{minipage}
  \hfill
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-print-not-constants.pdf}
    \caption{Most common bi-grams that appear in the code of print statements.}
    \label{fig:common-print-not-constants}
  \end{minipage}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{common-last-nodes.pdf}
  \caption{Most common nodes in the last statement of code cells.}
  \label{fig:common-last-nodes}
\end{figure}

We collected 1M last statements from 119K notebooks. Figure~\ref{fig:common-last-nodes} shows that the most frequent type of code written as the final statement in a code cell is a function call. We further analyse the functions calls. Figure~\ref{fig:common-last-modules}

% most common functions?

@@

Existence checks are a fundamental component of data preprocessing in ML, serving multiple critical purposes to ensure data integrity before analysis and modelling. These checks primarily focus on verifying the presence of necessary columns and the absence of missing values within those columns. Such validations are especially crucial after data preprocessing steps, where transformations might inadvertently introduce null values. These checks are also crucial for maintaining the accuracy and efficiency of data handling processes, ensuring that the datasets are ready for robust machine learning applications.

@@

Assertion \emph{A3} in Table~\ref{tab:maths-check} checks that the dimension differences between height of input and the filter, when divided by the stride, result in an integer, ensuring that convolution operations are dimensionally consistent. \emph{A25} monitors the standard deviation of outputs to decide the appropriateness of using batch normalisation, thus guiding the model optimization process based on statistical behaviour. \emph{A56} verifies the uniformity of certain computations, essential for maintaining consistency in state or parameter arrays over iterations. Similarly \emph{A64} checks for the symmetry of a matrix, a crucial property in many algorithms, particularly those involving covariance matrices and distance calculations.

When working with neural networks and other statistical models, it is imperative to ensure that mathematical properties are preserved throughout operations involving arrays and matrices. This rigour is captured through the assertions in Table~\ref{tab:maths-check} tha validate these properties after matrix manipulations. These assertions are not just precautionary but are vital for confirming that the underpinnings of ML algorithms align with expected mathematical principles, thereby ensuring the robustness and reliability of model computations and outcomes.

@@

% TODO: verify if this applies to recommender systems as well?
For example, \emph{A7} in Table~\ref{tab:model-perf-explicit} checks for the expected number of neighbours as an indirect measure of model performance in specific scenarios like recommendation systems or clustering. Meanwhile, direct comparisons of model outputs with expected results, as seen in \emph{A15} validate the model's predictive accuracy and reliability under operational conditions. Other assertions, such as \emph{A19}, ensure that the learned parameters align closely with theoretically or empirically derived values, further cementing the model’s statistical validity. Range checks like \emph{A38} confirm that the performance metrics such as \emph{AUC} and \emph{Gini coefficient} fall within acceptable boundaries, indicating good predictive performance without overfitting.

@@

Assertion \emph{A75} ensures that the regularization parameter \lstinline{reg} is either 'none' or 'l2'. This check is important for validating that the specified regularization method is among the acceptable options, thereby preventing potential misconfigurations that could lead to unexpected behaviour or suboptimal model performance.

@@

Type check assertions are essential for maintaining data integrity, ensuring compatibility between different components of the machine learning pipeline, and preventing errors and unexpected behaviour. By verifying the types of input features, models, data structures, and other objects, these assertions contribute to the overall reliability and robustness of the machine learning workflow.

@@

The visualisation in VA3 presents a two-dimensional scatter plot showing the decision boundary of a Support Vector Machine (SVM) model, with red and yellow points representing two different classes and a background colour gradient indicating the decision function of the SVM. This visualisation aids in understanding the model's ability to distinguish between classes and the distance of the samples from the decision boundary. The visualisations in VA28 display the training loss and development recall, respectively. The left graph shows a decrease in training loss over time, indicating learning progress, while the right graph shows development recall over time, highlighting improvements in the model's ability to correctly identify positive instances. The visualisations in VA11 depict key metrics for evaluating the performance of a machine learning model in text generation or machine translation. The left plot shows a steady increase in the development BLEU score over the training iterations, indicating improving quality of the model's output, while the right plot displays a steady decrease in training loss, suggesting effective learning and error minimization. In VA107, the scatter plot displays the predicted versus true values of Nitrogen Dioxide (NO2) levels, used to assess the model's performance in predicting continuous variables.
